\Large{Adversarial Training against Systematic Uncertainty}

### 2016-09-19 tex/paper.tex
save

\begin{proposition}\label{prop:1}
Let $\hat\theta_r = \arg \max_{\theta_r} E(\theta_f,
\theta_r)$ for fixed $\theta_f$. If $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$
for all $z \in {\cal Z}$, then $f$ is a pivotal quantity.
\end{proposition}

\begin{proof}
The optimal parameters $$\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} {\cal L}_r(\theta_f,
\theta_r)$$ are such that $p_{\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
By assumption, $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$,
and therefore $p(z|f(X;\theta_f)) = p(z)$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)}\\
                       &= p(f(X;\theta_f)),
\end{align*}
which holds for all $z \in {\cal Z}$ and implies that $f$ is a pivotal quantity.
\end{proof}

@@

For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r)$, in which case $p_{\theta_r}(z|f(X;\theta_f)) =
p(z|f(X;\theta_f))$ and ${\cal L}_r$ reduces to the entropy
$H(p_{Z|f(X;\theta_f)})$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = {\cal
L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)}).$$  In particular, we have the lower
bound $$H(p_{Y|X}) - H(p_Z) \leq {\cal L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)})$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ corresponds to
    the parameters of an optimal classifier, in which case the expected
    negative log-likelihood ${\cal L}_f$ of $Y|X$ reduces to its minimum
    value $H(p_{Y|X})$,
    \item all
   outcomes of $Z|f(X;\smash{\hat\theta}_f)$ are equally likely, in which case
   $p(z|f(X;\smash{\hat\theta}_f)) = p(z)$ for all $z \in {\cal
   Z}$ since we require a uniform prior by construction. Note that in the
   continuous case, the supremum of the differential entropy
   over continuous distributions on the same bounded support is also realized
   by the uniform distribution over that support.
\end{itemize}
Accordingly, the second condition implies that $p_{\theta_r}(z|f(X;\theta_f)) =
p(z)$ and therefore that at this point, because of Proposition~\ref{prop:1}, the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal quantity.

@@

Interestingly, let us emphasize that these results hold using only the (1D) output
of $f$ (in the case of binary classification) as input to the adversary. We
could similarly enforce an intermediate representation of the data to be
pivotal, e.g. as in \citep{ganin2014unsupervised}, but this is in fact not
necessary. \glnote{This also suggests we could use for $r$ a simpler, maybe
closed form, representation than a neural network (e.g. a GMM?).}

>>>>

\begin{proposition}\label{prop:1}
Let $\hat\theta_r = \arg \max_{\theta_r} E(\theta_f,
\theta_r)$ for fixed $\theta_f$. If $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$
for all $z \in {\cal Z}$, then $f$ is a pivotal quantity.
\end{proposition}

\begin{proof}
The optimal parameters $$\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} {\cal L}_r(\theta_f,
\theta_r)$$ are such that $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
By assumption, $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$,
and therefore $p(z|f(X;\theta_f)) = p(z)$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)}\\
                       &= p(f(X;\theta_f)),
\end{align*}
which holds for all $z \in {\cal Z}$ and implies that $f$ is a pivotal quantity.
\end{proof}

@@

For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r)$, in which case $p_{\hat\theta_r}(z|f(X;\theta_f)) =
p(z|f(X;\theta_f))$ and ${\cal L}_r$ reduces to the entropy
$H(p_{Z|f(X;\theta_f)})$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = {\cal
L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)}).$$  In particular, we have the lower
bound $$H(p_{Y|X}) - H(p_Z) \leq {\cal L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)})$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $H(p_{Y|X})$,

    \item $\smash{\hat\theta}_f$ maximimizes the entropy
    $H(p_{Z|f(X;\theta_f)})$, which happens when all outcomes of
    $Z|f(X;\smash{\hat\theta}_f)$ are equally likely and in which case
    $p(z|f(X;\smash{\hat\theta}_f)) = p(z)$ for all $z \in {\cal Z}$ since we
    require a uniform prior by construction. Note that in the continuous case,
    the supremum of the differential entropy over continuous distributions on
    the same bounded support is also realized by the uniform distribution over
    that support.
\end{itemize}
When the lower bound is active, we have $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$
because of the second condition, and because of Proposition~\ref{prop:1}, the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is therefore also a pivotal
quantity.

@@

Interestingly, let us emphasize that these results hold using only the (1D) output
of $f$ (in the case of binary classification) as input to the adversary. We
could similarly enforce an intermediate representation of the data to be
pivotal, e.g. as in \citep{ganin2014unsupervised}, but this is in fact not
necessary.
